Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2015-01-14 NewDesignDoc/DesignDocGLMM.tex
Completed first round of design doc edit/revamp

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
\end{align}
Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.

@@

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

@@

Recall $q_t$ is the number of random effects associated with variance component $\nu_t$. Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The model fitting function of \texttt{glmm} has made  $Z$ into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  is $q \times q$. Thus, each $E_t$  is $q \times q$.   $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal. We continue this process to complete the construction of $E_t, t=1,\ldots, T$.

\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
 Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is $s$ is defined such that, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$. The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$. Let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.  Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.  Let $\grave{f}(u)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate standard $t$ with $\zeta$ degrees of freedom. Then the importance sampling distribution is:
\begin{align}
 \tilde{f}(u) = p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
\end{align}
Then
\begin{align}
 \log \tilde{f}(u) = \log \left( p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
\end{align}

The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.  The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.

@@

 Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. Let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$. 
%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
Then we draw 
\begin{align}
u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
\end{align}
Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}.

@@

Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting hessian.

The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and hessian for the distribution of the random effects. The value is  evaluated by the C function \texttt{distRandGeneral}, described in section \ref{sec:Dgeneral}.

@@

@@

\section{Objective function: approximated log likelihood}\label{sec:objfun}
The objective function is optimized by \texttt{trust} within the the model fitting function.  In order to evaluate and maximize the approximated log likelihood, I need an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.  Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, \ref{eq:MCLAhessian}   for calculating these quantities.  This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.

Denote
\begin{align}
b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \label{eq:MCLA}
\end{align}

For computational stability, we'll set $a=max(b_k)$. Then the value of the MCLA is expressed as:\\
\begin{align}
\l_m (\theta) = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

@@

Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
\begin{align}
\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
\end{align}

Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
\begin{align}
&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
\end{align}
Everything in this has already been defined except:
\begin{align}
\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
\end{align}

\section{Summary of model}
Typing \texttt{summary(mod)}  gives more detailed info of the model. This is broken into two pieces (as is usually done for summaries):  \texttt{summary.glmm} and \texttt{print.summary.glmm}. We split this up because we have a lot of information in the summary that we do not necessarily want printed each time. When a user types \texttt{summary(mod)}, only the  basic information is automatically printed. More information can be found in the summary list.

The summary performs all calculations and its value is a list of the following:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix with the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column. All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
\item the evaluated Monte Carlo log likelihood along with its first and second derivative
\item maybe other things that I haven't thought of
\end{itemize}

A note on the standard errors: to calculate the standard errors, we take the MCLA Hessian matrix, invert it, take the diagonal elements, and square root them. It's possible that the Hessian will be noninvertible if it is close enough to singular that the computer thinks it's singular.  Then the standard errors will all be infinite. We also need to warn the user why this is happening.

Then \texttt{print.summary.glmm}  prints the following:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix similar to the output of summary.glm, which will be put through printCoefmat in order to get the significance stars that we're used to. We'll have the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column, the significance stars, and the significance legend.
\item maybe other things that I haven't thought of. I'll probably realize more once I get coding.
\end{itemize}

I think for the printCoefmat to work, the fixed effects matrix needs to look like
\begin{verbatim}
            Estimate Std. Error z value     Pr(>|z|)
(Intercept)        2       0.50       4 6.334248e-05
x1                 2       0.25       8 1.244192e-15
\end{verbatim}
(or at least it worked when I had these as the column names and didn't work before that).

Note that the summary and print.summary functions will be S3 generic functions.  This means is the user  types \texttt{summary(mod)} and \texttt{summary} is a generic function. R checks the class of the model \texttt{mod} and then automatically uses the \texttt{summary} and \texttt{print.summary} functions for that class of objects.

\section{PQL}\label{sec:pql}
 Before we get started on describing PQL, we need to change notation to avoid constrained optimization.  Recall that $D=\var(u)$ and is (for now) assumed to be diagonal. Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative. Using A rather than D enables unconstrained optimization.  If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
\begin{align*}
A= \sum_t E_t \sigma_t.
\end{align*}
Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix. PQL will estimate the components contained on the diagonal of A. Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).

In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.

There are two ways to do PQL, both of which are described in the vignette
 \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014). In either case, there is an inner optimization and an outer optimization. The inner optimization is well behaved while the other optimization is a little tougher. I will be using the method that is not quite PQL but is pretty close and is better behaved.  In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$. Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.

The {\bf inner} optimization will be done with the trust function in R. We elect to use trust because it requires two derivatives, which will make the optimization more precise. We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.

>>>>

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \label{eq:MCLAhessian}
\end{align}
Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and Hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.

@@

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the Hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

@@

Recall $q_t$ is the number of random effects associated with variance component $\nu_t$. Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The model fitting function of \texttt{glmm} has made  \texttt{z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to  count the  number of columns ($q_t$) for each design matrix in the list.
Since $D$  is $q \times q$, each $E_t$  is $q \times q$.   $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal. We continue this process to complete the construction of $E_t, t=1,\ldots, T$.

\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
 Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is $s$ is defined such that, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t},t=1,\ldots,T$.  Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$. The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$. Let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.  Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.  Let $\grave{f}(u)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate standard $t$ with $\zeta$ degrees of freedom. Then the importance sampling distribution is:
\begin{align}
 \tilde{f}(u) = p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
\end{align}

The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.  The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y,u)$.

@@

 Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. We randomly determine $m_1,m_2,m_3$ according to the proportions $p_1,p_2,p_3$. More specifically, we generate $m$ random numbers from a distribution that is uniform between $0$ and $1$. Then $m_1$ is the frequency with which the randomly generated numbers are between $0$ and $p_1$,  $m_2$ is the frequency with which the randomly generated numbers are between $p_1$ and $p_1+p_2$, and $m_3=m-p_1-p_2$.
%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.

Once $m_1, m_2$ and $m_3$ are determined, we draw 
\begin{align}
u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
\end{align}
Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}. We use R package \texttt{mvtnorm} to draw from a $t$ distribution.

@@

Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting Hessian.

The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and Hessian for the distribution of the random effects. The value is  evaluated by the C function \texttt{distRandGeneral}, described in section \ref{sec:Dgeneral}.

@@

Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int, and \texttt{logft} as a pointer.

@@

\section{Objective function: approximated log likelihood}\label{sec:objfun}
The objective function is optimized by \texttt{trust} within the the model fitting function.  In order to evaluate and maximize the approximated log likelihood, \texttt{trust} needs an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.  Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, and \ref{eq:MCLAhessian}   for calculating these quantities.  
%This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.

Denote
\begin{align}
b_k = \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k). \label{eq:MCLA}
\end{align}

For computational stability, we'll set $a=$max$(b_k)$. Then the value of the MCLA is expressed as:\\
\begin{align}
\l_m (\theta) = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

@@

Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
\begin{align}
\nabla l_m(\theta) &= \sum_{k=1}^m  \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)   \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]   \right) v_\theta(u_k,y)   \\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(u_k)  \right] v_\theta(u_k,y) \label{eq:MCLAgr}
\end{align}

Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
\begin{align}
&\nabla^2 l_m(\theta)=  \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)   v_\theta(u_k,y) \label{eq:MCLAh} \\ 
&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
    \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
\end{align}
Everything in this has already been defined except:
\begin{align}
\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
\end{align}

%HERE IS WHEN FTWIDDLE DEPENDS ON NU
%Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
%\begin{align}
%\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
%&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
%&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
%&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
%&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
%\end{align}
%
%Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
%\begin{align}
%&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
%   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
%&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
%   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
%   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
%\end{align}
%Everything in this has already been defined except:
%\begin{align}
%\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
%&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
%\end{align}
%
%
